---
title: GrafGen Benchmark

format:
  html:
    toc: true
    number-sections: true
    colorlinks: true

author:
  - name:   Difei Wang
    url:    
    email:  
    orcid:  
    note:   
    affiliations:
      - name:         Frederick National Laboratory for Cancer Research,
                      National Institutes of Health
        department:   Cancer Genomics Research Laboratory
        address:      9615 Shady Grove Rd
        city:         Rockville
        region:       MD
        country:      USA
        postal-code:  20850
        url:          
---

```{r}
# Install packages if needed (uncomment if you haven't installed them)
# install.packages(c("mclust", "aricode", "ggplot2", "ggalluvial"))

library(mclust)     # for adjustedRandIndex
library(aricode)    # for NMI
library(ggplot2)    # for plotting bar chart
library(ggalluvial) # for Sankey plot
```

# 1. Read CSV files

```{r}
grafgen    <- read.csv("data/grafgen_clusters.csv", header = TRUE)
fastbaps5  <- read.csv("data/fastBAPS_clusters_c5.csv", header = TRUE)
fastbaps7  <- read.csv("data/fastBAPS_clusters_c7.csv", header = TRUE)
fastbaps8  <- read.csv("data/fastBAPS_clusters_c8.csv", header = TRUE)
fastbaps13 <- read.csv("data/fastBAPS_clusters_c13.csv", header = TRUE)
poppunk    <- read.csv("data/poppunk_clusters.csv", header = TRUE)
#finestructure <- read.csv("finestructure_clusters.csv", header = TRUE)
FS <- read.csv("data/FS_clusters.csv", header = TRUE)
FSsub <- read.csv("data/FSsub_clusters.csv", header = TRUE)
DAP6 <- read.csv("data/DAP6_clusters.csv", header = TRUE)

# 2. Merge by Sample

```{r}
merged <- Reduce(function(x, y) merge(x, y, by = "Sample"),
                 list(grafgen, fastbaps5, fastbaps7,
                      fastbaps8, fastbaps13, FS, FSsub, DAP6, poppunk))

# Rename cluster columns for clarity
#colnames(merged) <- c("sample_id", "GrafGen", "fastBAPS", "PopPUNK",
#                      "fineSTRUCTURE")
colnames(merged) <- c("sample_id", "GrafGen", "fastBAPS_c5",
                      "fastBAPS_c7", "fastBAPS_c8", "fastBAPS_c13",
                      "FS", "FSsub", "DAP6", "PopPUNK")

# 3. Convert clusters to factors (important for ARI/NMI)
merged$GrafGen       <- as.factor(merged$GrafGen)
merged$fastBAPS_c5   <- as.factor(merged$fastBAPS_c5)
merged$fastBAPS_c7   <- as.factor(merged$fastBAPS_c7)
merged$fastBAPS_c8   <- as.factor(merged$fastBAPS_c8)
merged$fastBAPS_c13  <- as.factor(merged$fastBAPS_c13)
merged$PopPUNK       <- as.factor(merged$PopPUNK)
merged$FS <- as.factor(merged$FS)
merged$FSsub <- as.factor(merged$FSsub)
merged$DAP6 <- as.factor(merged$DAP6)
```

# 4. Compute ARI and NMI scores (each method vs GrafGen)

```{r}
ari_vals <- c(
  adjustedRandIndex(merged$GrafGen, merged$fastBAPS_c5),
  adjustedRandIndex(merged$GrafGen, merged$fastBAPS_c7),
  adjustedRandIndex(merged$GrafGen, merged$fastBAPS_c8),
  adjustedRandIndex(merged$GrafGen, merged$fastBAPS_c13),
  adjustedRandIndex(merged$GrafGen, merged$FS),
  adjustedRandIndex(merged$GrafGen, merged$FSsub),
  adjustedRandIndex(merged$GrafGen, merged$DAP6),
  adjustedRandIndex(merged$GrafGen, merged$PopPUNK)
)

nmi_vals <- c(
  NMI(merged$GrafGen, merged$fastBAPS_c5),
  NMI(merged$GrafGen, merged$fastBAPS_c7),
  NMI(merged$GrafGen, merged$fastBAPS_c8),
  NMI(merged$GrafGen, merged$fastBAPS_c13),
  NMI(merged$GrafGen, merged$FS),
  NMI(merged$GrafGen, merged$FSsub),
  NMI(merged$GrafGen, merged$DAP6),
  NMI(merged$GrafGen, merged$PopPUNK)
)

methods <- c("fastBAPS_c5", "fastBAPS_c7", "fastBAPS_c8",
             "fastBAPS_c13", "FS", "FSsub", "DAP6", "PopPUNK")
```

# 5. Prepare data frame for bar plot

```{r}
df_plot <- data.frame(
  Method = rep(methods, 2),
  Metric = rep(c("ARI", "NMI"), each = length(methods)),
  Score = c(ari_vals, nmi_vals)
)
```

# 6. Bar plot of ARI and NMI scores

```{r}
#pdf(file="results/ARI-NMI.barplot.pdf")
ggplot(df_plot, aes(x = Method, y = Score, fill = Metric)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.7),
           width = 0.6) +
  labs(title = "Clustering Similarity to GrafGen (pseudo ground truth)",
       y = "Score",
       x = "Method") +
  scale_fill_manual(values = c("ARI" = "#377eb8", "NMI" = "#4daf4a")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0, hjust=0.5))

#dev.off()
```

# 7. Prepare data for Sankey plot (all four methods)

```{r}
df_sankey <- merged[, c("sample_id", "GrafGen",
                                     "fastBAPS_c5",
                                     "fastBAPS_c7",
                                     "fastBAPS_c8",
                                     "fastBAPS_c13",
                                     "FS",
                                     "FSsub",
                                     "DAP6",
                                     "PopPUNK")]
head(df_sankey)
write.csv(df_sankey, file="results/df_sankey.csv")
```

# 8. Create Sankey plot showing flow across clusterings

```{r}
library(dplyr)

# Summarize counts for flows between GrafGen and fastBAPS (as example)
#flow_counts <- df_sankey %>%
#  group_by(GrafGen, fastBAPS) %>%
#  summarise(Freq = n()) %>%
#  ungroup()

#pdf(file="results/Sankey.pdf", height=16, width=24)
# 8. Create Sankey plot showing flow across clusterings
ggplot(df_sankey,
       aes(axis1 = GrafGen, axis2 = fastBAPS_c5, axis3 = fastBAPS_c7,
           axis4 = fastBAPS_c8, axis5 = fastBAPS_c13, axis6 = PopPUNK,
           axis7 = FS)) +
  geom_alluvium(aes(fill = GrafGen), width = 1/12) +
  geom_stratum(width = 1/12, fill = "gray90", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("GrafGen", "fastBAPS_c5", "fastBAPS_c7",
                              "fastBAPS_c8", "fastBAPS_c13", "PopPUNK", "FS"),
                   expand = c(.05, .05)) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) + 
  labs(title = "Sankey Plot: Cluster Membership Flow Across Methods",
       y = "Number of Samples")

# 8.1 Create Sankey plot showing flow across clusterings
ggplot(df_sankey,
       aes(axis1 = GrafGen,
           axis2 = fastBAPS_c5,
           axis3 = fastBAPS_c7,
           axis4 = fastBAPS_c8,
           axis5 = fastBAPS_c13,
           axis6 = FS,
           axis7 = FSsub,
           axis8 = DAP6)) +
  geom_alluvium(aes(fill = GrafGen), width = 1/12) +
  geom_stratum(width = 1/12, fill = "gray90", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)),
             size = 5, fontface = "bold") +
# geom_text(data = flow_counts,
#           aes(x = 1.5, y = Freq / 2 + cumsum(Freq) - Freq/2, label = Freq),
#           inherit.aes = FALSE,
#           size = 4) +
  scale_x_discrete(limits = c("GrafGen",
                              "fastBAPS_c5",
                              "fastBAPS_c7",
                              "fastBAPS_c8",
                              "fastBAPS_c13",
                              "FS",
                              "FSsub",
                              "DAP6"),
                   expand = c(.05, .05)) +
  theme_minimal(base_size = 18) +
  theme(
    axis.title = element_text(face = "bold", size = 18), # bold axis titles
    axis.text = element_text(face = "bold", size = 18), # bold axis text (ticks)
    plot.title = element_text(face = "bold", size = 22), # bold and bigger title
    axis.text.x = element_text(angle = 45, vjust = 0, hjust=1)) +
  labs(title = "Sankey Plot: Cluster Membership Flow Across Methods",
       x = "Methods",
       y = "Number of Samples")

## all methods
# 8.2  Create Sankey plot showing flow across clusterings
ggplot(df_sankey,
       aes(axis1 = GrafGen,
           axis2 = fastBAPS_c5,
           axis3 = fastBAPS_c7,
           axis4 = fastBAPS_c8,
           axis5 = fastBAPS_c13,
           axis6 = FS,
           axis7 = FSsub,
           axis8 = DAP6,
           axis9 = PopPUNK)) +
  geom_alluvium(aes(fill = GrafGen), width = 1/12) +
  geom_stratum(width = 1/12, fill = "gray90", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)),
             size = 5, fontface = "bold") +
# geom_text(data = flow_counts,
#           aes(x = 1.5, y = Freq / 2 + cumsum(Freq) - Freq/2, label = Freq),
#           inherit.aes = FALSE,
#           size = 4) +
  scale_x_discrete(limits = c("GrafGen",
                              "fastBAPS_c5",
                              "fastBAPS_c7",
                              "fastBAPS_c8",
                              "fastBAPS_c13",
                              "FS",
                              "FSsub",
                              "DAP6",
                              "PopPUNK"),
                   expand = c(.05, .05)) +
  theme_minimal(base_size = 18) +
  theme(
    axis.title = element_text(face = "bold", size = 18), # bold axis titles
    axis.text = element_text(face = "bold", size = 18), # bold axis text (ticks)
    plot.title = element_text(face = "bold", size = 22), # bold and bigger title
    axis.text.x = element_text(angle = 45, vjust = 0, hjust=1)) +
  labs(title = "Sankey Plot: Cluster Membership Flow Across Methods",
       x = "Methods",
       y = "Number of Samples")
#dev.off()
```

 9. Output ARI and NMI in a table and save to CSV

```{r}
results_table <- data.frame(
  Method = methods,
  ARI = round(ari_vals, 3),
  NMI = round(nmi_vals, 3)
)

print(results_table)

write.csv(results_table, "results/clustering_comparison_results.csv",
          row.names = FALSE)
```
